# Data Structures, Variable Types & Sampling NHANES

## Setup: Packages Used Here

```{r, message = FALSE}
knitr::opts_chunk$set(comment = NA)

library(NHANES)
library(janitor)
library(naniar)
library(tidyverse)

theme_set(theme_bw())
```

## Data require structure and context

**Descriptive statistics** are concerned with the presentation, organization and summary of data, as suggested in @Norman-Streiner. This includes various methods of organizing and graphing data to get an idea of what those data can tell us. 

As @Vittinghoff suggest, the nature of the measurement determines how best to describe it statistically, and the main distinction is between **numerical** and **categorical** variables. Even this is a little tricky - plenty of data can have values that look like numerical values, but are just numerals serving as labels. 

As @BockVD point out, the truly critical notion, of course, is that data values, no matter what kind, are useless without their contexts. The Five W's (Who, What [and in what units], When, Where, Why, and often How) are just as useful for establishing the context of data as they are in journalism. If you can't answer Who and What, in particular, you don't have any useful information.

In general, each row of a data frame corresponds to an individual (respondent, experimental unit, record, or observation) about whom some characteristics are gathered in columns (and these characteristics may be called variables, factors or data elements.) Every column / variable should have a name that indicates *what* it is measuring, and every row / observation should have a name that indicates *who* is being measured.

## Sampling Adults in NHANES {#newNHANES} 

In @sec-nhanes1, we spent some time with a sample from the National Health and Nutrition Examination. Now, by changing the value of the `set.seed` function which determines the starting place for the random sampling, and changing some other specifications, we'll generate a new sample describing 750 unique (distinct) adult subjects who completed the 2011-12 version of the survey when they were between the ages of 21 and 64.

### Creating a Temporary, Cleaner Data Frame

I'll start by describing the plan we will use to create a new data frame called `nh_temp` from which we will eventually build our final sample. In particular, let me lay out the steps I will use to create the `nh_temp` frame from the original `NHANES` data frame available in the R package called `NHANES`.

1. We'll **filter** the original NHANES data frame to include only the responses from the 2011-12 administration of the survey. This will cut the sample in half, from 10,000 rows to 5,000.

2. We'll then **filter** again to restrict the sample to adults whose age is at least 21 and also less than 65. I'll do this because I want to avoid problems with including both children and adults in my sample, and because I also want to focus on the population of people in the US who are usually covered by private insurance from their job, or by Medicaid insurance from the government, rather than those covered by Medicare.

3. As we discussed previously, what is listed in the NHANES data frame as `Gender` should be more correctly referred to as `Sex`. `Sex` is a biological feature of an individual, while `Gender` is a social construct. This is an important distinction, so I'll change the name of the variable.

4. We'll also rename three other variables, specifically we'll use `Race` to describe the `Race3` variable in the original NHANES data frame, as well as `SBP` to refer to the average systolic blood pressure, which is specified as `BPSysAve`, and `DBP` to refer to the average diastolic blood pressure, which is specified as `BPDiaAve`.

5. Having accomplished the previous four steps, we'll then *select* the variables we want to keep in the sample. (We use *select* for choosing variables or columns in the data frame, and *filter* for selecting subjects or rows.) The sixteen variables we will select are: ID, Sex, Age, Height, Weight, Race, Education, BMI, SBP, DBP, Pulse, PhysActive, Smoke100, SleepTrouble, MaritalStatus and HealthGen.

6. The original NHANES data frame includes some subjects (rows) multiple times in an effort to incorporate some of the sampling weights used in most NHANES analyses. For our purposes, though, we'd like to only include each subject one time. We use the `distinct()` function to limit the data frame to completely unique subjects (so that, for example, we don't wind up with two or more rows that have the same ID number.)

Here is the code I used to complete the six steps listed above and create the `nh_temp` data frame.

```{r build_nh_temp}
nh_temp <- NHANES |>
    filter(SurveyYr == "2011_12") |>
    filter(Age >= 21 & Age < 65) |>
    rename(Sex = Gender, Race = Race3, SBP = BPSysAve, DBP = BPDiaAve) |>
    select(ID, Sex, Age, Height, Weight, Race, Education, BMI, SBP, DBP, 
           Pulse, PhysActive, Smoke100, SleepTrouble, 
           MaritalStatus, HealthGen) |>
   distinct()
```

The resulting `nh_temp` data frame has `r nrow(nh_temp)` rows and `r ncol(nh_temp)` columns.

```{r}
nh_temp
```

### Sampling `nh_temp` to obtain our `nh_adult750` sample

Having established the `nh_temp` sampling frame, we now select a random sample of 750 adults from the `r nrow(nh_temp)` available responses. 

- We will use the `set.seed()` function in R to set a random numerical seed to ensure that if you redo this work, you will obtain the same sample.
    - Setting a seed is an important part of being able to replicate the work later when sampling is involved.
- Then we will use the `slice_sample()` function to actually draw the random sample, without replacement.
    - "Without replacement" means that once we've selected a particular subject, we won't select them again.

```{r}
set.seed(431002) 
# use set.seed to ensure that we all get the same random sample 

nh_adult750 <- slice_sample(nh_temp, n = 750, replace = F) 

nh_adult750
```

The `nh_adult750` data frame now includes `r nrow(nh_adult750)` rows (observations) on `r ncol(nh_adult750)` variables (columns). Essentially, we have `r ncol(nh_adult750)` pieces of information on each of `r nrow(nh_adult750)` adult NHANES subjects who were included in the 2011-12 panel.

### Summarizing the Data's Structure

We can identify the number of rows and columns in a data frame or tibble with the `dim` function.

```{r}
dim(nh_adult750)
```


The `str` function provides a lot of information about the structure of a data frame or tibble.

```{r}
str(nh_adult750)
```

To see the first few observations, use `head`, and to see the last few, try `tail`...

```{r}
tail(nh_adult750, 5) # shows the last five observations in the data set
```

### What are the variables?

We can use the `glimpse` function to get a short preview of the data.

```{r}
glimpse(nh_adult750)
```

The variables we have collected are described in the brief table below^[Descriptions are adapted from the ?NHANES help file. Remember that what NHANES lists as Gender is captured here as Sex, and similarly Race3, BPSysAve and BPDiaAve from NHANES are here listed as Race, SBP and DBP.].

Variable | Description | Sample Values
----------:  | ------------------------------------------ | -------------
ID           | a numerical code identifying the subject   | 68648, 67200
Sex          | sex of subject (2 levels)                  | female, male
Age          | age (years) at screening of subject        | 30, 35
Height       | height (in cm) at screening of subject     | 181.3, 180.2
Weight       | weight (in kg) at screening of subject    | 67.1, 86.6
Race         | reported race of subject (6 levels)        | White, Black
Education    | educational level of subject (5 levels)    | College Grad, High School
BMI          | body-mass index, in kg/m^2^                | 20.4, 26.7
SBP          | systolic blood pressure in mm Hg           | 103, 113
DBP          | diastolic blood pressure in mm Hg          | 59, 68
Pulse        | 60 second pulse rate in beats per minute   | 78, 70
PhysActive   | Moderate or vigorous-intensity sports?     | Yes, No
Smoke100     | Smoked at least 100 cigarettes lifetime?   | Yes, No
SleepTrouble | Told a doctor they have trouble sleeping?  | Yes, No
MaritalStatus | Marital Status    | Married, Divorced
HealthGen    | Self-report general health rating (5 levels) | Vgood, Fair

The levels for the multi-categorical variables are:

- **Race**: Mexican, Hispanic, White, Black, Asian, or Other.
- **Education**: 8th Grade, 9 - 11th Grade, High School, Some College, or College Grad.
- **MaritalStatus**: Married, Widowed, Divorced, Separated, NeverMarried or LivePartner (living with partner).
- **HealthGen**: Excellent, Vgood, Good, Fair or Poor.

Some details can be obtained using the `summary` function.

```{r}
summary(nh_adult750)
```

Note the appearance of `NA's` (indicating missing values) in some columns, and that some variables are summarized by a list of their (categorical) values (with counts) and some (quantitative/numeric) variables are summarized with a minimum, quartiles and means.

## Quantitative Variables

Variables recorded in numbers that we use as numbers are called **quantitative**. Familiar examples include incomes, heights, weights, ages, distances, times, and counts. All quantitative variables have measurement units, which tell you how the quantitative variable was measured. Without units (like miles per hour, angstroms, yen or degrees Celsius) the values of a quantitative variable have no meaning. 

- It does little good to be told the price of something if you don't know the currency being used. 

- You might be surprised to see someone whose age is 72 listed in a database on childhood diseases until you find out that age is measured in months.

- Often just seeking the units can reveal a variable whose definition is challenging - just how do we measure "friendliness", or "success," for example.

- Quantitative variables may also be classified by whether they are **continuous** or can only take on a **discrete** set of values. Continuous data may take on any value, within a defined range. Suppose we are measuring height. While height is really continuous, our measuring stick usually only lets us measure with a certain degree of precision. If our measurements are only trustworthy to the nearest centimeter with the ruler we have, we might describe them as discrete measures. But we could always get a more precise ruler. The measurement divisions we make in moving from a continuous concept to a discrete measurement are usually fairly arbitrary. Another way to think of this, if you enjoy music, is that, as suggested in @Norman-Streiner, a piano is a *discrete* instrument, but a violin is a *continuous* one, enabling finer distinctions between notes than the piano is capable of making. Sometimes the distinction between continuous and discrete is important, but usually, it's not.

    + The `nh_adult750` data includes several quantitative variables, specifically `Age`, `Height`, BMI`, `SBP`, `DBP` and `Pulse`.
    + We know these are quantitative because they have units: `Age` in years, `Height` in centimeters, `BMI` in kg/m^2^, the `BP` measurements in mm Hg, and `Pulse` in beats per minute.
    + Depending on the context, we would likely treat most of these as *discrete* given that are measurements are fairly crude (this is certainly true for `Age`, measured in years) although BMI is probably *continuous* in most settings, even though it is a function of two other measures (`Height` and `Weight`) which are rounded off to integer numbers of centimeters and kilograms, respectively.

- It is also possible to separate out quantitative variables into **ratio** variables or **interval** variables. An interval variable has equal distances between values, but the zero point is arbitrary. A ratio variable has equal intervals between values, and a meaningful zero point. For example, weight is an example of a ratio variable, while IQ is an example of an interval variable. We all know what zero weight is. An intelligence score like IQ is a different matter. We say that the average IQ is 100, but that's only by convention. We could just as easily have decided to add 400 to every IQ value and make the average 500 instead. Because IQ's intervals are equal, the difference between and IQ of 70 and an IQ of 80 is the same as the difference between 120 and 130. However, an IQ of 100 is not twice as high as an IQ of 50. The point is that if the zero point is artificial and movable, then the differences between numbers are meaningful but the ratios between them are not. On the other hand, most lab test values are ratio variables, as are physical characteristics like height and weight. A person who weighs 100 kg is twice as heavy as one who weighs 50 kg; even when we convert kg to pounds, this is still true. For the most part, we can treat and analyze interval or ratio variables the same way.

    + Each of the quantitative variables in our `nh_adult750` data can be thought of as ratio variables.

- Quantitative variables lend themselves to many of the summaries we will discuss, like means, quantiles, and our various measures of spread, like the standard deviation or inter-quartile range. They also have at least a chance to follow the Normal distribution.

### A look at BMI (Body-Mass Index)

The definition of BMI (*body-mass index*) for adult subjects (which is expressed in units of kg/m^2^) is:

$$
\mbox{Body Mass Index} = \frac{\mbox{weight in kg}}{(\mbox{height in meters})^2} = 703 \times \frac{\mbox{weight in pounds}}{(\mbox{height in inches})^2}
$$

> [BMI is essentially] ... a measure of a person's *thinness* or *thickness*... BMI was designed for use as a simple means of classifying average sedentary (physically inactive) populations, with an average body composition. For these individuals, the current value recommendations are as follow: a BMI from 18.5 up to 25 may indicate optimal weight, a BMI lower than 18.5 suggests the person is underweight, a number from 25 up to 30 may indicate the person is overweight, and a number from 30 upwards suggests the person is obese.
> 
> Wikipedia, https://en.wikipedia.org/wiki/Body_mass_index

## Qualitative (Categorical) Variables

**Qualitative** or categorical variables consist of names of categories. These names may be numerical, but the numbers (or names) are simply codes to identify the groups or categories into which the individuals are divided. Categorical variables with two categories, like yes or no, up or down, or, more generally, 1 and 0, are called **binary** variables. Those with more than two-categories are sometimes called **multi-categorical** variables. 

- When the categories included in a variable are merely names, and come in no particular order, we sometimes call them **nominal** variables. The most important summary of such a variable is usually a table of frequencies, and the mode becomes an important single summary, while the mean and median are essentially useless.

    + In the nh_adult750 data, `Race` is a nominal variable with multiple unordered categories. So is `MaritalStatus`.

- The alternative categorical variable (where order matters) is called **ordinal**, and includes variables that are sometimes thought of as falling right in between quantitative and qualitative variables. 

    + Examples of ordinal multi-categorical variables in the `nh_adult750` data include the Education and HealthGen variables.
    + Answers to questions like "How is your overall physical health?" with available responses Excellent, Very Good, Good, Fair or Poor, which are often coded as 1-5, certainly provide a perceived *order*, but a group of people with average health status 4 (Very Good) is not necessarily twice as healthy as a group with average health status of 2 (Fair). 

- Sometimes we treat the values from ordinal variables as sufficiently scaled to permit us to use quantitative approaches like means, quantiles, and standard deviations to summarize and model the results, and at other times, we'll treat ordinal variables as if they were nominal, with tables and percentages our primary tools.

- Note that all binary variables may be treated as ordinal, or nominal.

    + Binary variables in the `nh_adult750` data include `Sex`, `PhysActive`, `Smoke100`, `SleepTrouble`. Each can be thought of as either ordinal or nominal.

Lots of variables may be treated as either quantitative or qualitative, depending on how we use them. For instance, we usually think of age as a quantitative variable, but if we simply use age to make the distinction between "child" and "adult" then we are using it to describe categorical information. Just because your variable's values are numbers, don't assume that the information provided is quantitative.

## Counting Missing Values

The `summary()` command counts the number of missing observations in each variable, but sometimes you want considerably more information.

We can use some functions from the `naniar` package to learn useful things about the missing data in our `nh_adult750` sample.

The `miss_var_table` command provides a table of the number of variables with 0, 1, 2, up to n, missing values and the percentage of the total number of variables those variables make up.

```{r}
miss_var_table(nh_adult750)
```

So, for instance, we have `r miss_var_table(nh_temp) |> filter(n_miss_in_var == 0) |> select(n_vars)` variables with no missing data, and that constitutes `r miss_var_table(nh_temp) |> filter(n_miss_in_var == 0) |> select(pct_vars)`% of the 16 variables in our `nh_adult750` data.

The `miss_var_summary()` function tabulates the number, percent missing, and cumulative sum of missing of each variable in our data frame, in order of most to least missing values.

```{r}
miss_var_summary(nh_adult750)
```

So, for example, the `r `miss_var_summary(nh_temp) |> slice_head(n = 1) |> select(variable)` variable is the one missing more of our data than anything else within the `nh_adult750` data frame.

A graph of this information is available, as well.

```{r, warning = FALSE}
gg_miss_var(nh_adult750)
```

I'll note that there are also functions to count the number of missing observations by case (observation) rather than variable. For example, we can use `miss_case_table`.

```{r}
miss_case_table(nh_adult750)
```

Now we see that `r miss_case_table(nh_adult750) |> filter(n_miss_in_case == 0) |> select(n_cases)` observations, or `r miss_case_table(nh_adult750) |> filter(n_miss_in_case == 0) |> select(pct_cases)`% of all cases have no missing data.

We can use `miss_case_summary()` to identify cases with missing data, as well.

```{r}
miss_case_summary(nh_adult750)
```


## `nh_adults500cc`: A Sample of Complete Cases {#nh_cc} 

If we wanted a sample of exactly 750 subjects with complete data, we would have needed to add a step in the development of our `nh_temp` sampling frame to filter for complete cases.

```{r build_nh_temp2}
nh_temp2 <- NHANES |>
    filter(SurveyYr == "2011_12") |>
    filter(Age >= 21 & Age < 65) |>
    rename(Sex = Gender, Race = Race3, SBP = BPSysAve, DBP = BPDiaAve) |>
    select(ID, Sex, Age, Height, Weight, Race, Education, BMI, SBP, DBP, 
           Pulse, PhysActive, Smoke100, SleepTrouble, 
           MaritalStatus, HealthGen) |>
    distinct() |>
    na.omit()
```

Let's check that this new sampling frame has no missing data.

```{r}
miss_var_table(nh_temp2)
```

OK. Now, let's create a second sample, called `nh_adult500cc`, where now, we will select 500 adults with complete data on all of the variables of interest, and using a different random seed. The `cc` here stands for complete cases.

```{r}
set.seed(431003) 
# use set.seed to ensure that we all get the same random sample 

nh_adult500cc <- slice_sample(nh_temp2, n = 500, replace = F) 

nh_adult500cc
```

## Saving our Samples in `.Rds` files

We'll save the `nh_adult750` and `nh_adult500cc` samples to use in later parts of the notes. To do this, we'll save them as `.Rds` files, which will have some advantages for us later on.

```{r}
write_rds(nh_adult750, file = "data/nh_adult750.Rds")
write_rds(nh_adult500cc, file = "data/nh_adult500cc.Rds")
```

You will also find these `.Rds` files as part of the [`431-data` repository](https://github.com/THOMASELOVE/431-data) for the course.

Next, we'll load, explore and learn about some of the variables in these two samples.

